\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2024-01-05 forwards-and-backwards.tex
halfway through overfull hboxes

\SUBMITTED{June 14, 2023} % Edit.
\ACCEPTED{NA} % Edit.

@@

\newif\ifsubmission
\submissiontrue
% \submissionfalse
\ifsubmission
    % % stuff for submission
    \usepackage{lineno}
    % allow line numbers around math environments
    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
          \renewenvironment{#1}%
             {\linenomath\csname old#1\endcsname}%
             {\csname oldend#1\endcsname\endlinenomath}}% 
        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
          \patchAmsMathEnvironmentForLineno{#1}%
          \patchAmsMathEnvironmentForLineno{#1*}}%
        \AtBeginDocument{%
        \patchBothAmsMathEnvironmentsForLineno{equation}%
        \patchBothAmsMathEnvironmentsForLineno{align}%
        \patchBothAmsMathEnvironmentsForLineno{flalign}%
        \patchBothAmsMathEnvironmentsForLineno{alignat}%
        \patchBothAmsMathEnvironmentsForLineno{gather}%
        \patchBothAmsMathEnvironmentsForLineno{multline}%
        }
\else
    % for the nice version
\fi

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses}  % name of the review responses file

@@

An entirely analogous argument shows that if $G'''$ is bounded, then
the term $\epsilon_{\theta,N}(f, G, \eta)$ 
in~(\ref{generator prelimit}) will be $\bigO(\theta/N^2)$.

@@

Next we must choose the new level created at the birth event.
We would like an individual with level $u$ and at spatial position $x$
to produce offspring at $y$ at instantaneous rate
\begin{equation}
	\label{rate of birth}
	2 \Big(1 - \frac{u}{N}\Big) \theta \gamma(x, \eta) r(x + y, \eta).
\end{equation}
To do this
we will associate the point $(\tau, v, z, \kappa) \in \Pi_a$ with 
level $u + v \ell$,
where $\ell$ is chosen so that the rate of appearance of points in 
$\Pi_a$ with level below $N$, that is points with
$v \ell <N-u$, is given by~(\ref{rate of birth}).
Since the mean measure of $\Pi_a$ is Lebesgue measure
in the $t$ and $v$ directions, we must take
\begin{equation}
	\label{expression for ell}
    \ell(x, y, \eta)
    =
    \frac{N-u}{2 (1-u/N) \theta \gamma(x, \eta) r(x + y, \eta) } 
    =\frac{1}{2N^{-1} \theta \gamma(x, \eta) r(x + y, \eta) } 
    ,
\end{equation}
and, using this, the (potential) new level is
\begin{equation*}
    U_{a \concat j}(\tau)
    =
    U_a(\tau)
    +
    v \ell\big(X_a(\tau-), y(X_a(\tau-), z), \eta_{\tau-} \big) .
\end{equation*}
If $U_{a \concat j}(\tau) < N$,
the new individual labeled $a \concat j$ is produced,
and $\kappa$ determines which label, $a$ or $a\concat j$,
is associated with the new location,
so
\begin{align*}
    X_{a \concat j}(\tau)
    &=
    X_a(\tau-) + (1 - \kappa) y\big(X_a(\tau-), z\big) .
\end{align*}
On the other hand
if $U_{a \concat j}(\tau) \ge N$, then $X_a$ is unchanged and 
$X_{a \concat j}$ is undefined,
so
\begin{align} \label{eqn:X_line_of_descent}
    X_a(\tau)
    &=
    X_a(\tau-) + \kappa y(X_a(\tau-), z) \ind_{U_{a \concat j}(\tau) < N} .
\end{align}
Recall that the parental \emph{individual} always retains their spatial location,
so that $\kappa = 1$ corresponds to the parent being assigned a new level,
and our line of descent switching to the offspring.
Combining these observations, $X_a$, for $\tau_a \le t < \sigma_a$, solves the equation 
\begin{align*}
    X_a(t)
    &=
    X_a(\tau_a) + 
    \int_{[\tau_a, t) \times [0, \infty) \times \IR \times [0, 1]}
    y(X_a(\tau-), z)
    \kappa
    \ind_{
        U_a(\tau) + v \ell(X_a(\tau-), y(X_a(\tau-), z), \eta_{\tau-})<N
    }
    d\Pi_a(\tau, v, z, \kappa) .
\end{align*}

@@

First suppose that
assumption~\ref{control through r} of  
Lemma~\ref{lem:conditions on r} is satisfied. \revpoint{1}{28}
We write
\begin{align*}
    r(y,\eta) f(y) - r(x,\eta) f(x)
&=
    r(y,\eta) (f(y) - f(x))
    + (r(y,\eta) - r(x,\eta)) f(x) 
\\ &=
    r(y,\eta) \left(
        \sum_i (y-x)_i \partial_{x_i} f(x)
        + \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} f(z_1)
    \right)
\\ &\qquad {}
    + f(x) \left(
        \sum_i (y-x)_i \partial_{x_i} r(x,\eta)
        + \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} r(z_2,\eta)
    \right)
 \\ &=
    \left( r(x,\eta) + \sum_j (y-x)_j \partial_{x_j} r(z_3,\eta) \right)
    \left(
        \sum_i (y-x)_i \partial_{x_i} f(x)
    \right)
\\ &\qquad {}
    + r(y,\eta) 
        \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} f(z_1)
\\ &\qquad {}
    + f(x) \left(
        \sum_i (y-x)_i \partial_{x_i} r(x,\eta)
        + \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} r(z_2,\eta)
    \right) ,
\end{align*}
for some $z_i = \kappa_i x + (1-\kappa_i) y$.
Integrating this against $q(x, dy)$, we get that
\begin{align*}
& \bigg| \theta \int \left(
    r(y,\eta) f(y) - r(x,\eta) f(x)
\right) q_\theta(x, dy) \bigg|
\\ &\qquad \le
    \bigg| \sum_i \left( r(x,\eta) \partial_{x_i} f(x) + f(x) \partial_{x_i} r(x,\eta) \right)
    \theta \int (y-x)_i q_\theta(x, dy) \bigg|
\\ &\qquad \qquad {} +
    \bigg| f(x) 
    \theta \int \sum_{ij} \partial_{x_i x_j} r(z_2,\eta) (y-x)_i (y-x)_j q_\theta(x, dy) \bigg|
\\ &\qquad \qquad {} +
    \bigg| \theta \int \sum_{ij} (y-x)_i (y-x)_j 
        \left(
            \partial_{x_i} f(x) \partial_{x_j} r(z_3,\eta)
            + r(y,\eta) \partial_{x_ix_j} f(z_1)
        \right) q_\theta(x, dy) \bigg| .
\end{align*}
Since $q_\theta(x,dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$ are uniformly bounded,
$\theta \int (y-x)_i q_\theta(x,dy)$ is bounded as well.
Furthermore, a change of variables that diagonalizes $\covq(x)$
shows for any $g : \IR^d \to \IR^{d + d}$,
that if $C_g = \sup_y \sup_{\|z\| = 1} \sum_{ij} g(y)_{ij} z_i z_j$
and $\lambda_* = \sup_y \sup_{\|z\|=1} \sum_{ij} \covq(y)_{ij} z_i z_j$
then
$$ \theta \int \sum_{ij} g(y)_{ij} (y-x)_i (y-x)_j q_\theta(x, dy) \le C_g \lambda_* .  $$
Condition~\ref{control through r} 
gives uniform bounds on the derivatives of $r(x,\eta) = r(x,\rho_r*\eta(x))$
in this expression and so, provided $f$ also has uniformly bounded first and second derivatives,
we have a bound of the form
\begin{align*}
    |B^\theta_f| \leq K_1 + K_2 |f(x)| ,
\end{align*}
for suitable constants $K_1$, $K_2$ that depend only on the derivatives of $f$.

@@

\begin{proof}
    First note that by the proof of Lemma~\ref{lem:eta_f_bound},
    $\IE[\langle 1, \eta^N_t \rangle] \le \IE[\langle 1, \eta^N_0 \rangle] e^{Ct}$
    for some $C$ (independent of $N$).
    Now, let $M^{N*}_t(f) = \sup_{0 \le s \le t} M^N_t(f)$,
    and as before let $\langle M^N(f) \rangle_t$ be the angle bracket process of $M^N_t(f)$.
    The Burkholder-Davis-Gundy inequality says that there is a $K$ for which
    $\IE\left[ M^{N*}_t(1) \right] \le K \IE[\sqrt{[M^N(1)]_t}]$,
    where $[M^N(1)]_t$ is the quadratic variation of $M^N(1)$.
    Furthermore, as discussed by \citet{hernandez/jacka:2022},
    % for $q \le 1$ there exist universal constants $c_q$ such that
    % for any local martingale with $M_0=0$, $\IE[[M]^{q}_T] \le c_q \IE[\langle M \rangle^q_T]$
    the expectation of the quadratic variation of a local martingale
    is bounded by a (universal) constant multiple of the expectation of its angle bracket process
    (\cite{barlow/jacka/yor:1986}, Item (4.b'), Table 4.1, p. 162).
    Now, since $\sqrt{x} \le 1 + x$,
    in the notation of Lemma~\ref{lem:eta_f_bound}, there is a $C'$ such that
    \begin{align*}
        \IE\left[ M^{N*}_t(1) \right]
        &\le
        C'\left( 1 + \IE\left[ \left\langle M^N(1) \right\rangle_t \right] \right)
        \\ &=
        C'\left( 1 + \frac{\theta}{N} \IE\left[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\left\{
                \gamma(x, \eta^N_s)
                \int_{\IR^d} r(y, \eta^N_s) q_\theta(x, dy)
                + \mu_\theta(x, \eta^N_s)
            \right\}, \eta_s^N(dx)\Big\rangle
            ds
            \right] \right)
        \\ &=
        C'\Big( 1 + \IE\Big[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\Big\{
                \frac{2\theta}{N} \gamma(x, \eta^N_s) r(x, \eta^N_s)
        \\ &\qquad\qquad\qquad\qquad \qquad {}
		+ \frac{\gamma(x,\eta_s^N)}{N} 
		B^\theta_1(x, \eta^N_s)-\frac{1}{N}F(x,\eta^N_s)
            \Big\}, \eta_s^N(dx) \Big\rangle
            ds
            \Big] \Big) .
    \end{align*}

	We have not assumed that $F$ is bounded below, but to see that the 
	term involving $-F$ does not cause us problems, we rearrange
	equation~(\ref{eqn:eta_f_mgale_decomp}) 
	with $f=1$ to see that
	\begin{align}
		\label{integral of -F}
        \begin{split}
\IE\left[\int_0^t \Big\langle -F(x, \eta^N_s), \eta^N_s(dx) \Big\rangle ds \right]
        &= \IE[\langle 1, \eta^N_0\rangle] -\IE[\langle 1,\eta^N_t\rangle]
            \\&\qquad {}
	+ \IE\left[\int_0^t\Big\langle \gamma(x,\eta_s^N) B^\theta_1(x, \eta_s^N),\eta_s^N(dx)\Big\rangle ds\right],
        \end{split}
	\end{align}
which is bounded above since $\gamma (x,\eta)$ and $B^\theta_1(x,\eta)$ are both bounded and 
$\langle 1,\eta_t^N \rangle\geq 0$. 
    Since $\theta/N \to \alpha < \infty$,
    combining constants, we obtain that for some $C''$,
    \begin{align*}
        \IE\left[ M^{N*}_t(1) \right]
        &\le
        C' + C'' \IE[ \langle 1, \eta_0^N \rangle ] e^{tC} .
    \end{align*}
    Taking suprema and expectations on both sides of equation~\eqref{eqn:eta_f_mgale_decomp},
    then again using the fact that $\gamma(x, \eta) B^\theta_1(x, \eta) + F(x, \eta) \le C$,
    \begin{align*}
        \IE\left[\sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle \right]
        &\le
        \IE[\langle 1, \eta^N_0 \rangle]
        + \IE\left[
            \sup_{0 \le t \le T}
            \int_0^t 
	    \Big\langle\left\{
                \gamma(x, \eta^N_s)
                B^\theta_1(x, \eta^N_s)
                + F(x, \eta^N_s)
            \right\}, \eta^N_s(dx) \Big\rangle ds
        \right]
        \\ & \qquad\qquad {}
        + \IE[M^{N*}_t(1)] 
        \\
        &\le
        \IE[\langle 1, \eta^N_0 \rangle]
        + C \IE\left[
            \int_0^T \sup_{0 \le s \le t} \langle 1, \eta^N_s \rangle dt
        \right]
        + C' + C'' \IE[ \langle 1, \eta_0^N \rangle ] e^{tC} .
    \end{align*}
    Once again applying Gronwall's inequality,
    \begin{align*}
        \IE\left[\sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle \right]
        &\le
        C''' 
        \left(1+ \IE[\langle 1, \eta^N_0 \rangle]\right) e^{2TC} .
    \end{align*}
    For any $T$,
    the quantity on the right is bounded above by a constant $C(T)$ independent of $N$.
    As a result, for any $K > 0$,
    \begin{equation*}
    \limsup_{N \to \infty}
        \IP\left[ \sup_{0 \le s \le T} \langle 1, \eta^{N}_{s} \rangle \geq K \right]
        \leq
        \frac{C(T)}{K}.
    \end{equation*}
\end{proof}

@@

Similarly,
\begin{align*}
    &
 \big|   \langle M^{N}(f) \rangle_{\tau_N + t} 
        - \langle M^{N}(f) \rangle_{\tau_N} \big|
    \\ & \qquad =
  \Big|  \int_{\tau_N}^{\tau_N + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \int_{\IR^d} f^2(y) r(y, \eta^N_s) q_\theta(x, dy)
            +
            \mu_\theta(x, \eta^N_s) f^2(x)
        \right\}
        \eta^N_s(dx)
    ds \Big|
    \\ & \qquad =
   \Big| \int_{\tau_N}^{\tau_N + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \left(
                2 f^2(x) r(x, \eta^N_s)
                +
                B^\theta_{f^2}(x, \eta^N_s)
            \right)
            -
            f^2(x) \frac{F(x, \eta^N_s)}{ \theta}
        \right\}
        \eta^N_s(dx)
    ds\Big| ,
\end{align*}
and so using the fact that $\theta/N \to \alpha < \infty$,
an argument entirely analogous to 
that for \eqref{eqn:eta_projections_goal1}
yields a $\delta_2$ for which \eqref{eqn:eta_projections_goal2} is
satsified. Taking $\delta=\min\{\delta_1,\delta_2\}$, the result follows.

@@

We work on the time interval $[0,T]$.
We will require a lower bound on $\int \psi_t^\epsilon(x)\log \psi_t^\epsilon(x) dx$ 
which we record as a lemma.

@@

C\IE\Big[\Big\langle\int_0^t\sum_{j=2}^k\frac{\theta}{N\epsilon^d}
\Big(\frac{1}{(N\epsilon^d)^{j-2}}\Big)

@@

\bibliographystyle{amsplain}
\bibliography{refs}

@@

% \providecommand{\bysame}{\leavevmode\hbox to3em{\hrulefill}\thinspace}
% \providecommand{\MR}{\relax\ifhmode\unskip\space\fi MR }
% % \MRhref is called by the amsart/book/proc definition of \MR.
% \providecommand{\MRhref}[2]{%
%   \href{http://www.ams.org/mathscinet-getitem?mr=#1}{#2}
% }
% \providecommand{\href}[2]{#2}
% \begin{thebibliography}{10}
% 
% \bibitem{barlow/jacka/yor:1986}
% M.~T. Barlow, S.~D. Jacka, and M.~Yor, \emph{Inequalities for a pair of
%   processes stopped at a random time}, Proceedings of the London Mathematical
%   Society \textbf{s3-52} (1986), no.~1, 142--172.
% 
% \bibitem{barton:1979}
% N~H Barton, \emph{The dynamics of hybrid zones}, Heredity \textbf{43} (1979),
%   no.~3, 341--359.
% 
% \bibitem{berestycki2009nonlocal}
% Henri Berestycki, Gregoire Nadin, Benoit Perthame, and Lenya Ryzhik, \emph{The
%   non-local fisher-kpp equation: travelling waves and steady states},
%   Nonlinearity \textbf{22} (2009), no.~12, 2813--2844.
% 
% \bibitem{birzu/hallatschek/korolev:2017}
% Gabriel Birzu, Oskar Hallatschek, and Kirill~S. Korolev, \emph{Fluctuations
%   uncover a distinct class of travelling waves}, Proc. Natl. Acad. Sci. (USA)
%   \textbf{115} (2018), no.~16, 3645--3654.
% 

>>>>

\SUBMITTED{June 14, 2023}
\ACCEPTED{January 3, 2024}

@@

\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
    % % stuff for submission
    \usepackage{lineno}
    % allow line numbers around math environments
    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
          \renewenvironment{#1}%
             {\linenomath\csname old#1\endcsname}%
             {\csname oldend#1\endcsname\endlinenomath}}% 
        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
          \patchAmsMathEnvironmentForLineno{#1}%
          \patchAmsMathEnvironmentForLineno{#1*}}%
        \AtBeginDocument{%
        \patchBothAmsMathEnvironmentsForLineno{equation}%
        \patchBothAmsMathEnvironmentsForLineno{align}%
        \patchBothAmsMathEnvironmentsForLineno{flalign}%
        \patchBothAmsMathEnvironmentsForLineno{alignat}%
        \patchBothAmsMathEnvironmentsForLineno{gather}%
        \patchBothAmsMathEnvironmentsForLineno{multline}%
        }
\else
    % for the nice version
    \newcommand{\llabel}[1]{}
    \newcommand{\revpoint}[2]{}
    \newcommand{\includereviews}{}
\fi

% % Responses to reviews
% \input{review-response-commands}
% % set this to show line numbers and include responses to reviews or not
% \newif\ifreviewresponses
% % \reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
% \newcommand{\responsefile}{review-responses}  % name of the review responses file

@@

An analogous argument shows that if $G'''$ is bounded, then
the term $\epsilon_{\theta,N}(f, G, \eta)$ 
in~(\ref{generator prelimit}) will be $\bigO(\theta/N^2)$.

@@

Next we must choose the new level created at the birth event.
We would like an individual with level $u$ and at spatial position $x$
to produce offspring at $y$ at instantaneous rate
\begin{equation}
	\label{rate of birth}
	2 \Big(1 - \frac{u}{N}\Big) \theta \gamma(x, \eta) r(x + y, \eta).
\end{equation}
To do this
we will associate the point $(\tau, v, z, \kappa) \in \Pi_a$ with 
level $u + v \ell$,
where $\ell$ is chosen so that the rate of appearance of points in 
$\Pi_a$ with level below $N$, that is points with
$v \ell <N-u$, is given by~(\ref{rate of birth}).
Since the mean measure of $\Pi_a$ is Lebesgue measure
in the $t$ and $v$ directions, we must take
\begin{equation}
	\label{expression for ell}
    \ell(x, y, \eta)
    =
    \frac{N-u}{2 (1-u/N) \theta \gamma(x, \eta) r(x + y, \eta) } 
    =\frac{1}{2N^{-1} \theta \gamma(x, \eta) r(x + y, \eta) } 
    ,
\end{equation}
and, using this, the (potential) new level is
\begin{equation*}
    U_{a \concat j}(\tau)
    =
    U_a(\tau)
    +
    v \ell\big(X_a(\tau-), y(X_a(\tau-), z), \eta_{\tau-} \big) .
\end{equation*}
If $U_{a \concat j}(\tau) < N$,
the new individual labeled $a \concat j$ is produced,
and $\kappa$ determines which label, $a$ or $a\concat j$,
is associated with the new location,
so
\begin{align*}
    X_{a \concat j}(\tau)
    &=
    X_a(\tau-) + (1 - \kappa) y\big(X_a(\tau-), z\big) .
\end{align*}
On the other hand
if $U_{a \concat j}(\tau) \ge N$, then $X_a$ is unchanged and 
$X_{a \concat j}$ is undefined,
so
\begin{align} \label{eqn:X_line_of_descent}
    X_a(\tau)
    &=
    X_a(\tau-) + \kappa y(X_a(\tau-), z) \ind_{U_{a \concat j}(\tau) < N} .
\end{align}
Recall that the parental \emph{individual} always retains their spatial location,
so that $\kappa = 1$ corresponds to the parent being assigned a new level,
and our line of descent switching to the offspring.
Combining these observations, $X_a$, for $\tau_a \le t < \sigma_a$, solves the equation 
\begin{align*}
    &X_a(t)
    =
    X_a(\tau_a) \\
    &\qquad {} + 
    \int_{[\tau_a, t) \times [0, \infty) \times \IR \times [0, 1]}
    y(X_a(\tau-), z)
    \kappa
    \ind_{
        U_a(\tau) + v \ell(X_a(\tau-), y(X_a(\tau-), z), \eta_{\tau-})<N
    }
    d\Pi_a(\tau, v, z, \kappa) .
\end{align*}

@@

First suppose that
assumption~\ref{control through r} of  
Lemma~\ref{lem:conditions on r} is satisfied. \revpoint{1}{28}
We write
\begin{align*}
    &r(y,\eta) f(y) - r(x,\eta) f(x)
    \\
&\qquad\qquad
    =
    r(y,\eta) (f(y) - f(x))
    + (r(y,\eta) - r(x,\eta)) f(x) 
\\ &\qquad\qquad
    =
    r(y,\eta) \left(
        \sum_i (y-x)_i \partial_{x_i} f(x)
        + \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} f(z_1)
    \right)
\\ &\qquad\qquad
    \qquad {}
    + f(x) \left(
        \sum_i (y-x)_i \partial_{x_i} r(x,\eta)
        + \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} r(z_2,\eta)
    \right)
 \\ &\qquad\qquad
    =
    \left( r(x,\eta) + \sum_j (y-x)_j \partial_{x_j} r(z_3,\eta) \right)
    \left(
        \sum_i (y-x)_i \partial_{x_i} f(x)
    \right)
\\ &\qquad\qquad
    \qquad {}
    + r(y,\eta) 
        \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} f(z_1)
\\ &\qquad\qquad
    \qquad {}
    + f(x) \left(
        \sum_i (y-x)_i \partial_{x_i} r(x,\eta)
        + \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} r(z_2,\eta)
    \right) ,
\end{align*}
for some $z_i = \kappa_i x + (1-\kappa_i) y$.
Integrating this against $q(x, dy)$, we get that
\begin{align*}
& \bigg| \theta \int \left(
    r(y,\eta) f(y) - r(x,\eta) f(x)
\right) q_\theta(x, dy) \bigg|
\\ &\qquad \le
    \bigg| \sum_i \left( r(x,\eta) \partial_{x_i} f(x) + f(x) \partial_{x_i} r(x,\eta) \right)
    \theta \int (y-x)_i q_\theta(x, dy) \bigg|
\\ &\qquad \qquad {} +
    \bigg| f(x) 
    \theta \int \sum_{ij} \partial_{x_i x_j} r(z_2,\eta) (y-x)_i (y-x)_j q_\theta(x, dy) \bigg|
\\ &\qquad \qquad {} +
    \bigg| \theta \int \sum_{ij} (y-x)_i (y-x)_j 
        \left(
            \partial_{x_i} f(x) \partial_{x_j} r(z_3,\eta)
            + r(y,\eta) \partial_{x_ix_j} f(z_1)
        \right) q_\theta(x, dy) \bigg| .
\end{align*}
Since $q_\theta(x,dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$ are uniformly bounded,
$\theta \int (y-x)_i q_\theta(x,dy)$ is bounded as well.
Furthermore, a change of variables that diagonalizes $\covq(x)$
shows for any $g : \IR^d \to \IR^{d + d}$,
that if $C_g = \sup_y \sup_{\|z\| = 1} \sum_{ij} g(y)_{ij} z_i z_j$
and $\lambda_* = \sup_y \sup_{\|z\|=1} \sum_{ij} \covq(y)_{ij} z_i z_j$
then
$$ \theta \int \sum_{ij} g(y)_{ij} (y-x)_i (y-x)_j q_\theta(x, dy) \le C_g \lambda_* .  $$
Condition~\ref{control through r} 
gives uniform bounds on the derivatives of $r(x,\eta) = r(x,\rho_r*\eta(x))$
in this expression and so, provided $f$ also has uniformly bounded first and second derivatives,
we have a bound of the form
\begin{align*}
    |B^\theta_f| \leq K_1 + K_2 |f(x)| ,
\end{align*}
for suitable constants $K_1$, $K_2$ that depend only on the derivatives of $f$.

@@

\begin{proof}
    First note that by the proof of Lemma~\ref{lem:eta_f_bound},
    $\IE[\langle 1, \eta^N_t \rangle] \le \IE[\langle 1, \eta^N_0 \rangle] e^{Ct}$
    for some $C$ (independent of $N$).
    Now, let $M^{N*}_t(f) = \sup_{0 \le s \le t} M^N_t(f)$,
    and as before let $\langle M^N(f) \rangle_t$ be the angle bracket process of $M^N_t(f)$.
    The Burkholder-Davis-Gundy inequality says that there is a $K$ for which
    $\IE\left[ M^{N*}_t(1) \right] \le K \IE[\sqrt{[M^N(1)]_t}]$,
    where $[M^N(1)]_t$ is the quadratic variation of $M^N(1)$.
    Furthermore, as discussed by \citet{hernandez/jacka:2022},
    % for $q \le 1$ there exist universal constants $c_q$ such that
    % for any local martingale with $M_0=0$, $\IE[[M]^{q}_T] \le c_q \IE[\langle M \rangle^q_T]$
    the expectation of the quadratic variation of a local martingale
    is bounded by a (universal) constant multiple of the expectation of its angle bracket process
    (\cite{barlow/jacka/yor:1986}, Item (4.b'), Table 4.1, p. 162).
    Now, since $\sqrt{x} \le 1 + x$,
    in the notation of Lemma~\ref{lem:eta_f_bound}, there is a $C'$ such that
    \begin{align*}
        &\IE\left[ M^{N*}_t(1) \right]
        \le
        C'\left( 1 + \IE\left[ \left\langle M^N(1) \right\rangle_t \right] \right)
        \\ &\hspace{3em}=
        C'\left( 1 + \frac{\theta}{N} \IE\left[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\left\{
                \gamma(x, \eta^N_s)
                \int_{\IR^d} r(y, \eta^N_s) q_\theta(x, dy)
                + \mu_\theta(x, \eta^N_s)
            \right\}, \eta_s^N(dx)\Big\rangle
            ds
            \right] \right)
        \\ &\hspace{3em}=
        C'\Big( 1 + \IE\Big[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\Big\{
                \frac{2\theta}{N} \gamma(x, \eta^N_s) r(x, \eta^N_s)
        \\ &\qquad\qquad\qquad\qquad \qquad {}
		+ \frac{\gamma(x,\eta_s^N)}{N} 
		B^\theta_1(x, \eta^N_s)-\frac{1}{N}F(x,\eta^N_s)
            \Big\}, \eta_s^N(dx) \Big\rangle
            ds
            \Big] \Big) .
    \end{align*}

	We have not assumed that $F$ is bounded below, but to see that the 
	term involving $-F$ does not cause us problems, we rearrange
	equation~(\ref{eqn:eta_f_mgale_decomp}) 
	with $f=1$ to see that
	\begin{align}
		\label{integral of -F}
        \begin{split}
\IE\left[\int_0^t \Big\langle -F(x, \eta^N_s), \eta^N_s(dx) \Big\rangle ds \right]
        &= \IE[\langle 1, \eta^N_0\rangle] -\IE[\langle 1,\eta^N_t\rangle]
            \\&\qquad {}
	+ \IE\left[\int_0^t\Big\langle \gamma(x,\eta_s^N) B^\theta_1(x, \eta_s^N),\eta_s^N(dx)\Big\rangle ds\right],
        \end{split}
	\end{align}
which is bounded above since $\gamma (x,\eta)$ and $B^\theta_1(x,\eta)$ are both bounded and 
$\langle 1,\eta_t^N \rangle\geq 0$. 
    Since $\theta/N \to \alpha < \infty$,
    combining constants, we obtain that for some $C''$,
    \begin{align*}
        \IE\left[ M^{N*}_t(1) \right]
        &\le
        C' + C'' \IE[ \langle 1, \eta_0^N \rangle ] e^{tC} .
    \end{align*}
    Taking suprema and expectations on both sides of equation~\eqref{eqn:eta_f_mgale_decomp},
    then again using the fact that $\gamma(x, \eta) B^\theta_1(x, \eta) + F(x, \eta) \le C$,
    \begin{align*}
        \IE\left[\sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle \right]
        &\le
        \IE[\langle 1, \eta^N_0 \rangle]
        + \IE[M^{N*}_t(1)] 
        \\ & \qquad\qquad {}
        + \IE\left[
            \sup_{0 \le t \le T}
            \int_0^t 
	    \Big\langle\left\{
                \gamma(x, \eta^N_s)
                B^\theta_1(x, \eta^N_s)
                + F(x, \eta^N_s)
            \right\}, \eta^N_s(dx) \Big\rangle ds
        \right]
        \\
        &\le
        \IE[\langle 1, \eta^N_0 \rangle]
        + C \IE\left[
            \int_0^T \sup_{0 \le s \le t} \langle 1, \eta^N_s \rangle dt
        \right]
        + C' + C'' \IE[ \langle 1, \eta_0^N \rangle ] e^{tC} .
    \end{align*}
    Once again applying Gronwall's inequality,
    \begin{align*}
        \IE\left[\sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle \right]
        &\le
        C''' 
        \left(1+ \IE[\langle 1, \eta^N_0 \rangle]\right) e^{2TC} .
    \end{align*}
    For any $T$,
    the quantity on the right is bounded above by a constant $C(T)$ independent of $N$.
    As a result, for any $K > 0$,
    \begin{equation*}
    \limsup_{N \to \infty}
        \IP\left[ \sup_{0 \le s \le T} \langle 1, \eta^{N}_{s} \rangle \geq K \right]
        \leq
        \frac{C(T)}{K}.
    \end{equation*}
\end{proof}

@@

Similarly,
\begin{align*}
    &
 \big|   \langle M^{N}(f) \rangle_{\tau_N + t} 
        - \langle M^{N}(f) \rangle_{\tau_N} \big|
    \\ & \hspace{0.2em} =
  \Big|  \int_{\tau_N}^{\tau_N + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \int_{\IR^d} f^2(y) r(y, \eta^N_s) q_\theta(x, dy)
            +
            \mu_\theta(x, \eta^N_s) f^2(x)
        \right\}
        \eta^N_s(dx)
    ds \Big|
    \\ & \hspace{0.2em} =
   \Big| \int_{\tau_N}^{\tau_N + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \left(
                2 f^2(x) r(x, \eta^N_s)
                +
                B^\theta_{f^2}(x, \eta^N_s)
            \right)
            -
            f^2(x) \frac{F(x, \eta^N_s)}{ \theta}
        \right\}
        \eta^N_s(dx)
    ds\Big| ,
\end{align*}
and so using the fact that $\theta/N \to \alpha < \infty$,
an argument entirely analogous to 
that for \eqref{eqn:eta_projections_goal1}
yields a $\delta_2$ for which \eqref{eqn:eta_projections_goal2} is
satsified. Taking $\delta=\min\{\delta_1,\delta_2\}$, the result follows.

@@

We work on the time interval $[0,T]$, and
will require a lower bound on $\int \psi_t^\epsilon\log \psi_t^\epsilon dx$ 
which we record as a lemma.

@@

C\IE\Big[\Big\langle\int_0^t\sum_{j=2}^k
    \frac{\theta}{N\epsilon^d (N\epsilon^d)^{j-2}}

@@

\bibliographystyle{amsplain}
% \bibliography{refs}

@@

\providecommand{\bysame}{\leavevmode\hbox to3em{\hrulefill}\thinspace}
\providecommand{\MR}{\relax\ifhmode\unskip\space\fi MR }
% \MRhref is called by the amsart/book/proc definition of \MR.
\providecommand{\MRhref}[2]{%
  \href{http://www.ams.org/mathscinet-getitem?mr=#1}{#2}
}
\providecommand{\href}[2]{#2}
\begin{thebibliography}{10}

\bibitem{adler2018interspecific}
Peter~B. Adler, Andrew Kleinhesselink, Giles Hooker, Joshua~B. Taylor, Brittany
  Teller, and Stephen~P. Ellner, \emph{Weak interspecific interactions in a
  sagebrush steppe? {Conflicting} evidence from observations and experiments},
  Ecology \textbf{99} (2018), no.~7, 1621--1632.

\bibitem{barlow/jacka/yor:1986}
M.~T. Barlow, S.~D. Jacka, and M.~Yor, \emph{Inequalities for a pair of
  processes stopped at a random time}, Proceedings of the London Mathematical
  Society \textbf{s3-52} (1986), no.~1, 142--172.

\bibitem{barton:1979}
N~H Barton, \emph{The dynamics of hybrid zones}, Heredity \textbf{43} (1979),
  no.~3, 341--359.

\bibitem{berestycki2009nonlocal}
Henri Berestycki, Gregoire Nadin, Benoit Perthame, and Lenya Ryzhik, \emph{The
  non-local {Fisher}-{KPP} equation: travelling waves and steady states},
  Nonlinearity \textbf{22} (2009), no.~12, 2813--2844.
